Dependent Object Types

### 2012-08-05 article.tex
revise dot-one

\boxfig{fig:dot-mem-exp}{The DOT Calculus : Membership and Expansion}{
    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]

>>>>

\end{multicols}
} % END dot-one

The DOT calculus is a simple system of dependent
object-types. \figref{fig:dot-one} gives its syntax, reduction rules,
and type assignment rules.

\subsection*{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
%The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
%introduce a variable which is already bound in the base-set.
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

\subsection*{Syntax}

There are four alphabets: Variable names $x$, $y$, $z$ are freely
alpha-renamable. They occur as parameters of lambda abstractions, as
binders for objects created by \verb@new@-expressions, and as self
references in refinements. Value labels $l$ denote fields in objects,
which are bound to values at run-time. Similarly, method labels $m$
denote methods in objects. Type labels $L$ denote type members of
objects. Type labels are further separated into labels for abstract
types $L_a$ and labels for classes $L_c$. It is assumed that in each
program every class label $L_c$ is declared at most once.

We assume that the label alphabets $l$, $m$ and $L$ are finite. This is
not a restriction in practice, because one can include in these 
alphabets every label occurring in a given program.

The terms $t$ in DOT consist of variables $x$, $y$, $z$, field
selections $t.l$, method invocations $t.m(t)$ and object creation
expressions $\new y c {t'}$ where $c$ is a constructor $T_c \ldefs{\seq{l
    = v}\;\seq{m(x) = t}}$. The latter binds a variable $y$ to a new
instance of type $T_c$ with fields $\seq l$ initialized to values
$\seq v$ and methods $\seq m$ initialized to methods of one parameter
$\seq{x}$ and body $\seq{t}$.  The scope of $y$ extends through the term
${t'}$.

Two subclasses of terms are values $v$, which consist of just
variables, and paths $v$ which consist of just variables and field
selections.

The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
\begin{itemize}
\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
         other declarations in the same type by selecting from $z$.
\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
\item[-] A top type $\Top$, which corresponds to an empty object.
\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
\end{itemize}
A subset of types $T_c$ are called {\em concrete types}. These are type selections
$p.L_c$ of class labels,
the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.

There are only three forms of declarations in DOT, which are both part
of refinement types.  A value declaration $\ldecl l T$ introduces a
field with type $T$.  A method declaration $\mdecl m S U$ introduces a
method with parameter of type $S$ and result of type $U$. A type
declaration $\Ldecl L S U$ introduces a type member $L$ with a lower
bound type $S$ and an upper bound type $U$. There are no type aliases,
but a type alias can be simulated by a type declaration $\Ldecl L T T$
where the lower bound and the upper bound are the same type $T$.

Every field or type label can be declared only once in a set of
declarations $\seq D$. A set of declarations can hence be seen as a map from
labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
declarations are defined in ~\figref{fig:dot-decls}.

\subsection*{Reduction rules}

Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
terms $t$ and stores $s$, where stores map variables to constructors.
There are three main reduction rules: Rule (\textsc{msel}) rewrites a
method invocation $y {m_i} v$ by retrieving the corresponding method
definition from the store, and performing a substitution of the
argument for the parameter in the body. Rule (\textsc{sel}) rewrites a
field selection $x.l$ by retrieving the corresponding value from the
store. Rule (\textsc{new}) rewrites an object creation $\new x c t$ by
placing the binding of variable $x$ to constructor $c$ in the store
and continuing with term $t$.  These reduction rules can be applied
anywhere in a term where the hole $[\,]$ of an evaluation context $e$
can be situated.

\subsection*{Type assignment rules}

The last part of \figref{fig:dot-one} presents rules for type
assignment.

Rules (\textsc{sel}) and (\textsc{msel}) type field selections and
method invocations by means of an auxiliary membership relation $\ni$,
which determines whether a given term contains a given declaration as
one of its members. The membership relation is defined in
\figref{fig:dot-two} and is further explained below.

The last rule, (\textsc{new}), assigns types to object creation
expressions. It is the most complex of DOT's typing rules.  To
type-check an object creation $\new y {T_c \ldefs {\seq{l =
      v}\;\seq{m(x) = t}}} t'$, one verifies first that the type $T_c$
is well-formed (see \figref{fig:dot-wf} for a definition of
well-formedness).  One then determines the set of all declarations
that this type carries, using the expansion relation $\expand$ defined
in \figref{fig:dot-two}.  Every type declaration $\Ldecl L S U$ in
this set must be realizable, i.e.\ its lower bound $S$ must be a
subtype of its upper bound $U$.  Every field declaration $\ldecl l V$
in this set must have a corresponding initializing value of $v$ of
type $V$.  These checks are made in an environment which is extended
by the binding $y: T_c$. In particular this allows field values that
recurse on ``self'' by referring to the bound variable $x$. Similarly,
every method declaration $\mdecl m T W$ must have a corresponding
initializing method definition $m(x) = t$. The parameter type $T$ must
be $\wfe$ (well-formed and expanding; see \figref{fig:dot-wf}), and
the body $t$ must type check to $W$ in an environment extended by the
bindings $y : T_c$ and $x : T$.

Instead of adding a separate subsumption rule, subtyping is expressed
by preconditions in rules (\textsc{msel}) and (\textsc{new}).

\boxfig{fig:dot-two}{The DOT Calculus : Membership and Expansion}{
    \begin{multicols}{2}[\judgement{Membership}{\fbox{$\Gamma \ts t \ni D$}}]
